Determining the existance and direction of causality in the face of measurement error

### 2016-05-08 docs/manuscript.Rmd
udpated

fig_nums(
	"d_relationship_figure", 
	caption="Plots depicting the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$. $d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_O)$) on the x-axis. Higher measurement precision of $y$ ($cor(y,y_O)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.",
	display=FALSE
)

@@

where $\beta_x$ is the true effect of the exposure on the outcome. In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$. The 4th condition is necessary for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$. When measurement error is introduced we can show using basic covariance properties (Appendix 1) that

@@

### Inferring the existence of causality

When selecting an instrument $g$ that has a direct influence on $x$ and $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$. Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as

@@

### Inferring the direction of causality

If we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that the Steiger test will obtain the correct result. This is because the value $d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal association between $x$ and $y$ is $\rho_{x, y} < 1$. But it can be shown that in the presence of measurement error, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is important to note that under certain conditions when measurement error is present the Steiger test could make erroneous inference about causal direction because the value of $d$ is not restricted to being greater than 0. However, using the MR approach there are two potential advantages over the CIT. First, we can conduct a formal test for the direction of causality. Second the correct direction of causality can be inferred when measurement error in the exposure variable is present, as long as it is lower than the product of the measurement error in the outcome and the causal correlation between the exposure and the outcome. `r cfig("d_relationship_figure")` shows that in most cases, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.

We performed simulations to explore the performance of this approach in comparison to CIT. `r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, in a large number of cases the MR analysis infers the wrong direction of causality at a much lower rate than the CIT. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data, and unfortunately there are no solutions that work perfectly in all scenarios. In the face of measurement error it is evident that causal inference drawn through mediation analysis will be difficult to interpret, and this is a serious problem considering that it is often impossible to estimate the extent of measurement error for most experimental designs.

In the epidemiological literature this is relatively well understood and our initial analysis simply confirms that for widely used methods such as CIT they are indeed liable to the same issues as standard mediation based analysis. Specifically, we show that as measurement error in the exposure variable increases, CIT is likely to infer the wrong direction of causality. We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it increases the significance of the incorrect inference.

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of mediation based methods. Mendelian randomisation is robust in the face of measurement error and, given that the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance to the trait of interest. But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers two main advantages. First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

Mediation based network approaches are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust solutions alternatives for these more complicated problems.

>>>>

fig_nums(
	"d_relationship_figure", 
	caption="Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative. When $d$ is negative the Steiger test is liable to infer the wrong direction of causality. $d$ is plotted on the y axis, with measurement precision of $x$ ($cor(x,x_O)$) on the x-axis. Higher measurement precision of $y$ ($cor(y,y_O)$) is depicted by lighter coloured lines, and columns of boxes are for different true values of $cor(x,y)$. Shaded regions show the parameter space where $d$ is negative. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.",
	display=FALSE
)

@@

where $\beta_x$ is the true effect of the exposure on the outcome. In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$. The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$. When measurement error is introduced it can be shown using basic covariance properties (Appendix 1) that

@@

### Using MR to infer the existence of causality

When selecting an instrument $g$ that has a direct influence on $x$ where $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$. Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as

@@

### Using MR to infer the direction of causality

If we do not know whether the SNP $g$ has a direct influence on $x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$ is a valid instrument for $x$ then we expect that $\rho_{g,x}$ to be greater than $\rho_{g,y}$ because the association of $g$ with $y$ is the proportion of the true causal association between $x$ and $y$ scaled by the association between $g$ and $x$. We can use the Steiger test of correlated correlations to formally test the difference in magnitude of $\rho_{g,x}$ and $\rho_{g,y}$, where the variable with the largest correlation on $g$ is deemed to be the variable which is imposing the causal effect.

This is a simple method to orient the direction of causality in an MR analysis when the underlying biology of the SNP is not fully understood. But it can be shown that in the presence of measurement imprecision, $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), where $d$ is a metric of the degree to which the Steiger test is liable to provide the wrong estimate (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction). Thus it is important to note that under certain conditions when measurement imprecision is present the Steiger test could make erroneous inference about the causal direction because the value of $d$ is not restricted to being greater than 0. However `r cfig("d_relationship_figure")` shows that when there is no measurement error in $y$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $x$. It also shows that in most cases where $y$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the condition of $d > 0$ is satisfied.

We performed simulations to explore the performance of this approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect. `r cfig("cit_mr_comparison_figure")` shows that when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, the CIT is in general much more fallable to reporting a robust causal association for the wrong direction of causality. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data, and using genetic instruments as surrogates for randomisation is a useful approach. MR and the CIT employ genetic instruments in different ways, and unfortunately neither offer solutions that work perfectly in all scenarios. In the face of measurement error it is evident that causal inference drawn through mediation based methods such as the CIT analysis typically fare worse than MR and will be difficult to interpret, and this is a serious problem considering that it is often impossible to estimate the extent of measurement error present in observational data.

In the epidemiological literature this is relatively well understood and our initial analysis simply confirms that for widely used methods such as CIT they are indeed liable to the same issues as standard mediation based analysis. Specifically, we show that as measurement error in the exposure variable increases, CIT is likely to infer the wrong direction of causality. We also demonstrate that, though seemingly unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation. Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest. But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers two main advantages. First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

Mediation based network approaches are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012] network MR [@Burgess2015] and mediation through MR [@Varbo2015] may offer more robust alternatives for these more complicated problems.
